Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-07-09 POPL/paper.tex
single column
Added comments

% -*- fill-column: 80; -*-
\documentclass[sigconf,anonymous,review,screen]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows, fit}

@@

\section{A Syntax-Independent Notion of Syntax}
\label{sec:intro}
The theory of formal languages and parsing is one of the oldest and
most thoroughly developed areas of theoretical computer science. A
prominent topic in the 1950, 60s, and 70s led to a series of remarkable
developments: Chomsky's hierarchy of grammar formalisms \cite{chomThreeModels1956}, practical
algorithms for parsing of regular and context-free grammars and
variants \cite{KNUTH1965607,Earley1970}, as well as implementations of practical tools for the
generation of efficient parsers \cite{Johnsonyacc}.

@@

for every string $w \in \String$. Intuitively, we read a morphism of grammars as
a \emph{parse transformer} : $f$ translates a $G$-parse of $w$ into an $H$-parse
of $w$. The notion of isomorphism induced by
this categorical structure encodes that two grammars are equivalent if there is
a bijective translation of parses --- precisely Chomsky's notion of \emph{strong equivalence} of
grammars \cite{chom1963}.

@@

The category-theoretic framework can be further used to describe and compare
different notions of formal grammar. For instance, Chomskyan generative
grammars, semi-Thue systems, Montague grammars, and so on are all distinct syntactic
presentations of the same underlying idea of an abstract specification for
parsing. Each of these is an instance of a general \emph{notion of formal
  grammar} --- a category paired with a functor into
$\Set^{\String}$. That is, they may all be interpreted in our category of grammars. In this sense, $\Set^{\String}$ provides a universal setting to reason about parsing.
The slice 2-category over $\Set^{\String}$ then provides the appropriate expressivity to prove equivalences between grammar formalism, such as the equivalence between semi-Thue systems and Chomsky's unrestricted grammars.

@@

This paper begins in \cref{sec:categorify} by giving a categorical interpretation of Kleene algebra. However, Kleene algebras are primarily concerned with the regular languages. With this in mind, we further abstract our categorical model from the connectives of regular grammars to those of context-free grammars and beyond. From there, in \cref{sec:tt} we provide a type theory to construct and reason about such formal grammars. In \cref{sec:semantics} we give interpretations of the syntax as grammars and as languages, and prove a canonicity theorem for the type theory. Finally, in \cref{sec:automata} we give a presentation of automata as grammars. With this view in mind, we may encode parsing algorithms and execute them entirely internal to our formalism. Our contributions are then:

\section{Categorification of Formal Grammars}
\label{sec:categorify}
\steven{Idk if this section has the right title, but I wanted to change it if we are adding ``Chomsky categories''}

Kleene algebras are an important tool in the theory of regular
languages. More broadly, they serve as a theoretical substrate to
studying various kinds of formal languages. Formally, they are a tuple
$(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is a set, $+$ and $\cdot$
are binary operations over $A$, $(-)^*$ is a function over $A$, and
$1$ and $0$ are constants. These structures satisfy the axioms depicted
in Figure~\ref{fig:axioms}.

@@

This result feels similar in spirit to the definition of action
algebras, which are algebras where the product also has adjoint
operations which results in the Kleene star being more easily
axiomatized \cite{KOZEN1994366}.

We are now ready to prove that our concept of formal grammars fits
nicely within our categorical framework. We start by presenting a
well-known construction from presheaf categories.

@@

Much like in the posetal case, the abstract structure of a Kleene
category is expressive enough to synthetically reason about formal
languages. A significant difference between them is that while Kleene
algebras can reason about language containment, Kleene categories can
reason about parsing as well.

@@

\paragraph{Beyond Simple Types}

Though Kleene categories are expressive enough to reason about
concepts that are outside of reach of Kleene algebras, their
simply-typed nature makes them not so expressive from a type theory
point of view. This is limiting because type theories are tested
syntactic frameworks for manipulating complicated categories
structures while avoiding some issues common in category theory, such
as coherence issues.

With this in mind, we want to design a dependent type theory that
builds on top of Kleene categories with the goal of internalizing
various reasoning principles for languages and their parsers. This
leads us to the abstract notion of model we are interested in
capturing with our type theory: a \emph{Chomsky category}.

\begin{definition}
  A Chomsky category is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives,
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

@@

Further, the syntactic category of the theory in \cref{sec:tt} is manifestly a Chomsky category.
\steven{This is true, right? That's like the entire point of having the syntax}

@@

Omission of the structural rules of a deductive system, such as in
linear logic \cite{GIRARD19871}, offers precise control over how a value is used
in a derivation. Namely, linear logic omits the weakening and contraction rules
to ensure that every value in context exactly once. This control enables
\emph{resource-sensitive} reasoning, where we may treat a resource as
\emph{consumed} after usage. This viewpoint is amenable to parsing applications,
as we may treat characters of a string finite resources that are consumed at
parse-time. That is, once a particular substring has been parsed, a parser need not go
over it again. One may then envision a linear type system where the types comprise
formal grammars generated over some alphabet $\Sigma$, and the type constructors
correspond precisely to inductive constructions on grammars --- such as
conjunction, disjunction, concatenation, etc.

Unfortunately, programming in a linear term language is often cumbersome and
unintuitive for users. Code in such a language can become unnecessarily verbose
when using Girard encodings to translate high-level reasoning into low-level
linear terms. To alleviate this concern, in 1995 Benton et al.\ gave categorical
models of a logic that synthesized linear and non-linear logic
\cite{bentonMixedLinearNonlinear1995} --- appropriately referred to as a
\emph{linear-non-linear} (LNL) system. This work is simply typed, so the
boundary between linear and non-linear subtheories is entirely characterized via
a monoidal adjunction $F \dashv G$ between linear terms and non-linear terms.

@@

Inside of an LNL system, linearity may be thought of as an option that users can
choose to deliberately invoke at deliberate points in their developments in an
otherwise intuitionistic system. However, if we are wishing to treat parsers as
linear terms over input strings, the non-linear fragment of an LNL theory does
not really assist in the development of parsers. It is instead the case that
parsers may benefit from a \emph{dependence} on non-linear terms.

@@

Krishnaswami et al.\ equipped LNL systems with a restricted sense of dependent
types \cite{krishnaswami_integrating_2015}. When defining dependent types in
this setting dependence on linear terms is disallowed, but either linear or
non-linear terms are allowed to depend on non-linear terms.

@@

@@

\section{Canonicity}

Let $\cdot \vdash A$ be a closed linear type. Then there are
two obvious notions of what constitutes a ``parse'' of a string w
according to the grammar $A$:
\begin{enumerate}
\item On the one hand we have the set-theoretic semantics just
  defined, $\llbracket A \rrbracket \cdot w$
\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
  \lceil w \rceil \vdash e : G$.
\end{enumerate}
It is not difficult to see that at least for the ``purely positive''
formulae (those featuring only the positive connectives
$0,+,I,\otimes,\mu, \overline\Sigma,c$) that
every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
that the nodes of the tree correspond precisely to the introduction
forms of the type. However it is far less obvious that \emph{every}
linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
sequence of introduction forms since proofs can include elimination
forms as well. To show that this is indeed the case we give a
\emph{canonicity} result for the calculus: that the parses for .

@@

\section{Automata as Grammars}
\label{sec:automata}
Classically, formal language theory is closely related to the study of automata.
In the Chomsky hierarchy, each language class is associated to a class of
automata that serve as recognizers. Internal to the type theory of \cref{sec:tt},
we can characterize these language classes syntactically; moreover, we
demonstrate the equivalence of these language classes to its associated automata class as a
proof term within our logic.

@@

As we will see in \cref{subsubsec:tm}, we may very quickly encode arbitrary
recursively enumerable languages --- since we can simulate the
tape of Turing machine in the dependent fragment of the language. We reflect
here on this expressive power to motivate why we put stringent syntactic
restrictions on particular classes of automata. The type theory we present truly
is general enough to easily capture all of the Chomsky hierarchy.
To begin, let's focus on the simplest languages and their corresponding
automata --- regular expressions and finite automata.

@@

>>>>

% -*- fill-column: 80; -*-
\documentclass[sigconf,anonymous,review,screen,acmsmall]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows, fit}

@@

\section{Introduction}
\label{sec:intro}

The theory of formal languages and parsing is one of the oldest and
most thoroughly developed areas of theoretical computer science. A
prominent topic in the 1950, 60s, and 70s led to a series of remarkable
developments: Chomsky's hierarchy of grammar formalisms \cite{chomThreeModels1956}, practical
algorithms for parsing of regular and context-free grammars and
variants \cite{KNUTH1965607,Earley1970}, as well as implementations of practical tools for the
generation of efficient parsers \cite{Johnsonyacc}.

\paragraph{Contributions.} This paper begins in \cref{sec:categorify} by giving a categorical interpretation of Kleene algebra. However, Kleene algebras are primarily concerned with the regular languages. With this in mind, we further abstract our categorical model from the connectives of regular grammars to those of context-free grammars and beyond. From there, in \cref{sec:tt} we provide a type theory to construct and reason about such formal grammars. In \cref{sec:semantics} we give interpretations of the syntax as grammars and as languages, and prove a canonicity theorem for the type theory. Finally, in \cref{sec:automata} we give a presentation of automata as grammars. With this view in mind, we may encode parsing algorithms and execute them entirely internal to our formalism. Our contributions are then:

\pedro{Something good is coming up with a name to the type theory... This makes it easier to
refer to it throughout the paper. Spitballing some suggestions GTT (grammar type theory --- conflicts with Max's Gradual Type Theory), KTT (Kleene type theory)...}

\section{A Syntax-Independent Notion of Syntax}

@@

for every string $w \in \String$. Intuitively, we read a morphism of grammars as
a \emph{parse transformer}: $f$ translates a $G$-parse of $w$ into an $H$-parse
of $w$. The notion of isomorphism induced by
this categorical structure encodes that two grammars are equivalent if there is
a bijective translation of parses --- precisely Chomsky's notion of \emph{strong equivalence} of
grammars \cite{chom1963}.

\pedro{Written like this, it seems like this correspondence iso-strong
  equivalence is a well-known fact, but this is not an obvious
  observation.}

@@

The category-theoretic framework can be further used to describe and compare
different notions of formal grammar. For instance, Chomskyan generative
grammars, semi-Thue systems, Montague grammars, and so on are all distinct syntactic
presentations of the same underlying idea of an abstract specification for
parsing. Each of these is an instance of a general \emph{notion of formal
  grammar} --- a category paired with a functor into
$\Set^{\String}$. That is, they may all be interpreted in our category of grammars. In this sense, $\Set^{\String}$ provides a universal setting to reason about parsing.
The slice 2-category over $\Set^{\String}$ then provides the appropriate expressivity to prove equivalences between grammar formalism, such as the equivalence between semi-Thue systems and Chomsky's unrestricted grammars.
\pedro{why?}

@@

\section{Categorifying Formal Grammars}
\label{sec:categorify}
\steven{Idk if this section has the right title, but I wanted to change it if we are adding ``Chomsky categories''}

Kleene algebras are an important tool in the theory of regular
languages. More broadly, through various extensions, they serve as a
theoretical substrate to studying different kinds of formal languages.

We posit that by seeing Kleene algebra through the lens of category
theory, we gain some insight into the type-theoretic essence of formal
grammars. We begin this section by providing a ``simply-typed''
version of a categorification of Kleene algebra, which we call Kleene
category, and then show define a more expressive class of categories,
which we call Chomsky category, that will serve as the categorical
semantics to our type theory.

\subsection{Kleene categories}
A Kleene algebra is a tuple $(A, +, \cdot, (-)^*, 1, 0)$, where $A$ is
a set, $+$ and $\cdot$ are binary operations over $A$, $(-)^*$ is a
function over $A$, and $1$ and $0$ are constants. These structures
satisfy the axioms depicted in Figure~\ref{fig:axioms}.

@@

Something surprising about this lemma is that it provides an alternative
perspective on the observation that if a Kleene algebra has an
residuation operation, also called action algebra \cite{kozen1994action},
then the Kleene star admits a simpler axiomatization.

Since we want Kleene categories to generalize our notion of formal
grammars as presheaves $\String \to \Set$, we prove that they do
indeed form a Kleene category. We start by presenting a well-known
construction from presheaf categories.

@@

Much like in the posetal case, the abstract structure of a Kleene
category is expressive enough to synthetically reason about formal
languages. A significant difference between them is that while Kleene
algebras can reason about language containment, Kleene categories can
reason about parsing as well, as illustrated by the presheaf model
above.

@@

\subsection{Beyond Simple Types}

Though Kleene categories are expressive enough to reason about
concepts that are outside of reach of Kleene algebras, their
simply-typed nature makes them not so expressive from a type theoretic
point of view. This is limiting because type theories are sucessful
syntactic frameworks for manipulating complicated categorical
structures while avoiding some issues common in category theory, such
as coherence issues.

With this in mind, we want to design a categorical semantics that
builds on top of Kleene categories with the goal of extending them
with dependent types and making them capable reasoning about languages
and their parsers. This leads us to the abstract notion of model we
are interested in capturing with our type theory: a \emph{Chomsky
category}.

\pa{We should probably define some of the words in this definition}
\begin{definition}
  A Chomsky category is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives,
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

@@

Further, the syntactic category of the theory in \cref{sec:tt} is manifestly a Chomsky category.
\steven{This is true, right? That's like the entire point of having the syntax}
\pedro{This is likely true, but if we explicitly say so, this warants a proof. I think that
if we don't say anything about the syntactic category, reviewers won't mind.}

@@

Omission of the structural rules of a deductive system, such as in
linear logic \cite{GIRARD19871}, offers precise control over how a value is used
in a derivation. Namely, linear logic omits the weakening and contraction rules
to ensure that every value in context is used exactly once. This control enables
\emph{resource-sensitive} reasoning, where we may treat a resource as
\emph{consumed} after usage. This viewpoint is amenable to parsing applications,
as we may treat characters of a string as finite resources that are consumed at
parse-time. That is, once a particular substring has been parsed, a parser need not go
over it again. One may then envision a linear type system where the types comprise
formal grammars generated over some alphabet $\Sigma$, and the type constructors
correspond precisely to inductive constructions on grammars --- such as
conjunction, disjunction, concatenation, etc.

Unfortunately, programming in a linear term language is often
cumbersome and unintuitive for users. Code in such a language can
become unnecessarily verbose when using Girard encodings to translate
high-level reasoning into low-level linear terms. To alleviate this
concern, in 1995 Benton et al.\ proposed an alternative categorical
semantics of linear logic based on adjoint interactions between linear
and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
appropriately referred to as a \emph{linear-non-linear} (LNL)
system. This work is simply typed, so the boundary between linear and
non-linear subtheories is entirely characterized via a monoidal
adjunction $F \dashv G$ between linear terms and non-linear terms.

@@

Inside of an LNL system, linearity may be thought of as an option that users can
choose to deliberately invoke at deliberate points in their developments in an
otherwise intuitionistic system. However, if we are wishing to treat parsers as
linear terms over input strings, the non-linear fragment of an LNL theory does
not really assist in the development of parsers. It is instead the case that
parsers may benefit from a \emph{dependence} on non-linear terms.
\pedro{This last sentence could use an example. Is there a bite-sized, well-known
example we could mention here?}

@@

Krishnaswami et al.\ equipped LNL systems with a restricted sense of dependent
types \cite{krishnaswami_integrating_2015}. When defining dependent types in
this setting, dependence on linear terms is disallowed, but either linear or
non-linear terms are allowed to depend on non-linear terms.

@@

\pedro{I would argue that both paragraphs above should either be put
  somewhere else or deleted. I think that we should definitely mention
  Neel's LNL system, but as it stands, it feels disconnected from what
  came before and what comes next. A suggestion: the mention of Aho's
  indexed grammars could be the example we use illustrating the
  benefits of dependent grammars.}

@@

\pedro{After reading this section after the Kleene category one, I
  think that it makes much more sense to switch their order. It's
  going to be easier to justify providing a categorical semantics once
  we have a ``target'' of what we want to interpret.}

\section{Canonicity}
\pedro{It could be cool to come up with a catchy title to this section.
Something like ``Parse Canonicity''}

\pedro{Added the ``initial context'' paragraph below}

Canonicity is an important metatheoretic theorem in the type theory
literature.  It provides insight on the normal forms of terms and,
therefore, on its computational aspects. Frequently, proving
canonicity for boolean types, i.e. every closed term of type bool
reduces to either true or false, is enough to justifying that the type
theory being studied is well-behaved. In our case, however, since we
want to connect our type theory to parsers, we must provide a more
detailed account of canonicity.

If $\cdot \vdash A$ is a closed linear type then there are
two obvious notions of what constitutes a ``parse'' of a string w
according to the grammar $A$:
\begin{enumerate}
\item On the one hand we have the set-theoretic semantics just
  defined, $\llbracket A \rrbracket \cdot w$
\item On the other hand, we can view the string $w = c_1c_2\cdots$ as
  a linear context $\lceil w \rceil = x_1:c_1,x_2:c_2,\ldots$ and
  define a parse to be a $\beta\eta$-equivalence class of linear terms $\cdot;
  \lceil w \rceil \vdash e : G$.
\end{enumerate}
It is not difficult to see that at least for the ``purely positive''
formulae (those featuring only the positive connectives
$0,+,I,\otimes,\mu, \overline\Sigma,c$) that
every element $t \in \llbracket A \rrbracket w$ is a kind of tree and
that the nodes of the tree correspond precisely to the introduction
forms of the type. However it is far less obvious that \emph{every}
linear term $\lceil w \rceil \vdash p : \phi$ is equal to some
sequence of introduction forms since proofs can include elimination
forms as well. To show that this is indeed the case we give a
\emph{canonicity} result for the calculus: that the parses for .

@@

\section{Automata as Grammars}
\label{sec:automata}
\pedro{The switch from the canonicity proof to this section feels a bit ``bumpy''.
  Since they are at least technically related, it might be a good idea to say something
  like ``The canonicity theorem from the previous section gives a tight connection
  between terms in our type theory and the theory of formal languages. In this section
we further explore these connections by using the fact that, clasically ...''}

Classically, formal language theory is closely related to the study of automata.
In the Chomsky hierarchy, each language class is associated to a class of
automata that serve as recognizers. Internal to the type theory of \cref{sec:tt},
we can characterize these language classes syntactically; moreover, we
demonstrate the equivalence of these language classes to its associated automata class as a
proof term within our logic.

@@

As we will see in \cref{subsubsec:tm}, we may very quickly encode arbitrary
recursively enumerable languages --- since we can simulate the
tape of Turing machine in the dependent fragment of the language.We reflect
here on this expressive power to motivate why we put stringent syntactic
restrictions on particular classes of automata. The type theory we present truly
is general enough to easily capture all of the Chomsky hierarchy.
To begin, let's focus on the simplest languages and their corresponding
automata --- regular expressions and finite automata.
\pedro{The connection dependent types-Turing machines seems too important to only
  be mentioned here. It should at least indirectly motivate the type theory being
dependently typed.}

@@

\pedro{I like the overall structure of this section! Showing how the type theory
can encode and reason about different kinds of languages is very instructive!}
